Query Optimization for Dynamic Imputation

### 2017-05-15 paper/conclusion.tex
Update discussion of regression tree error results

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational algebra and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

>>>>

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational algebra and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered (\lowsmapealphazero{}-\highsmapealphaone{}\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

### 2017-05-15 paper/experiments.tex
Update discussion of regression tree error results

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline.
This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.

To calculate SMAPE for $\alpha=0$, $\alpha=0.5$, and $\alpha=1$, 
we compute tuple-wise absolute percentage deviations within each iteration of each query, and
average this value over all iterations. We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the baseline is low --- between \lowsmapealphazero{} and
\highsmapealphaoneexacs{} percent --- indicating that on-the-fly imputation produces results similar to the baseline. Notably, the results with largest error correspond
most closely to one existing approach of dropping all \nullv{}s.

\review{In \Cref{table:smape}, we report results using the reference chained-equation
decision trees model. We also experiment with mean-value imputation and hot
deck imputation, as outlined in \Cref{sec:imputation}. One approach to
mean-value imputation, to facilitate non-blocking operation within the iterator
model, is to estimate the column mean from base table histograms. In this case,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low no matter
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.}

@@

\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
\caption{\review{
    \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
    For clarity, we have removed from the plot plans that provide nearly indistinguishable trade-offs in either dimension.
}}
\label{fig:pareto-frontiers}
\end{figure}

>>>>

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for \ProjectName{}'s query results compared to the baseline, using the reference chained-equation decision trees model.
This measures the error introduced by on-the-fly imputation as compared to full imputation on the base tables.
This comparison is the relevant one in practice --- an analyst would only be considering the trade-offs in imputation quality and time in the first place for dirty input data, for which ground truth is unavailable.

To calculate SMAPE for each $\alpha$,
we compute tuple-wise absolute percentage deviations within each iteration of each query, and
average this value over all iterations.
Relative errors of query results, when optimizing for quality
($\alpha=0$), are low --- between \lowsmapealphazero{} and
\highsmapealphazero{} --- indicating that on-the-fly imputation produces
results similar to the baseline. When optimizing for performance ($\alpha=1$),
relative errors can be modestly higher --- up to \highsmapealphaone{} in the
worst case, as this approach most closely corresponds to dropping all
\nullv{}s. Thus, it is important to recognize the tradeoffs between quality and
performance for the specific use case.

\review{
We also experiment with mean-value imputation and hot
deck imputation, as outlined in \Cref{sec:imputation}. One approach to
mean-value imputation, to facilitate non-blocking operation within the iterator
model, is to estimate the column mean from base table histograms. In this case,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low no matter
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.
}
}

@@

\review{
\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/pareto_frontiers_plot.png}
\caption{\review{
    \ProjectName{}'s final planner Pareto frontier for \Cref{q1} and \Cref{q6}. 
    For clarity, we have removed from the plot plans that provide nearly indistinguishable trade-offs in either dimension.
}
\label{fig:pareto-frontiers}
}
\end{figure}

### 2017-05-15 paper/introduction.tex
Update discussion of regression tree error results

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

% Scope
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run
their queries without an explicit base-table imputation step.
% SAM -- cut the following -- could consider moving later but feels out of place here
%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches.
These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazero{} times as fast as it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaone{}
percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

\subsection{Contributions}
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run
their queries without an explicit base-table imputation step. To do so, it leverages a number of contributions
to minimize imputation time while producing comparable results to traditional approaches.

These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-05-15 paper/main.tex
Add qualitative results from other imputation methods
Update discussion of regression tree error results

% results that can be used in any section
\newcommand{\demorows}{10175}
\newcommand{\labexrows}{9813}
\newcommand{\acsbaseresultminutes}{355 minutes}
\newcommand{\acsbaseresulthours}{just under 6 hours}
\newcommand{\acsimputedbzeroresult}{4 seconds}
\newcommand{\acsimputedboneresult}{1 second}
\newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
\newcommand{\highxalphazeroexacs}{140} % the largest speed up, comparing a=0 to base table imputation 
\newcommand{\highxalphaoneexacs}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation 
\newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
\newcommand{\highsmapealphazero}{6} % the largest SMAPE difference, comparing a=0 to base table imputation
\newcommand{\highsmapealphaoneexacs}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
%\newcommand{\highxalphaone}{16000}
\newcommand{\nullv}{NULL}

@@

\newcommand{\planningmaxtime}{34} % TODO

@@

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets.
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
\end{abstract}

>>>>

% results that can be used in any section
\newcommand{\demorows}{10175}
\newcommand{\labexrows}{9813}
\newcommand{\acsbaseresultminutes}{355 minutes}
\newcommand{\acsbaseresulthours}{just under 6 hours}
\newcommand{\acsimputedbzeroresult}{4 seconds}
\newcommand{\acsimputedboneresult}{1 second}
\newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
\newcommand{\highxalphazero}{140} % the largest speed up, comparing a=0 to base table imputation 
\newcommand{\highxalphaone}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation 
\newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
\newcommand{\highsmapealphazero}{6} % the largest SMAPE difference, comparing a=0 to base table imputation
\newcommand{\highsmapealphaone}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
%\newcommand{\highxalphaone}{16000}
\newcommand{\nullv}{NULL}

@@

\newcommand{\planningmaxtime}{\todo{XXX}} % {34} % TODO

@@

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets.
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazero{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaone{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
\end{abstract}
